---
title: "Using GRID3 Data"
author: "Alina Game"
date: "10 November 2019"
output:
  html_document: default
---

This exercise provides an example of how we can use GRID3 data within an application. 

> This code is available via GitHub at www.github/GRID3/data-example

# Exercise Overview

Using geospatial analysis in R, this demo will show an assessment of health facility coverage for maternal health using the GRID3 population data.

In this exercise we are going to:

1. Load some spatial data
2. Subset the data to focus on points of interest
3. Aggregate dataset using some basic geospatial techniques
4. Try and identify connections between datasets

# Loading Packages

For this demo, we will be using the **sp** package along with **raster** to provide most of our spatial functions. The **dismo** and **spatialEco** package provide additional functions which we will use for some of our spatial analysis.

```{r setup, include=FALSE}
require(sp)
require(raster)
require(dismo)
require(spatialEco)
require(ggplot2)
```

# Loading Datasets

We have three datasets used in this example:

<!--- Can we include references to the datasets. Are these publicly available? ---> 

1. **Health facility locations**: 
2. **Ward boundaries**
3. **Gridded population (women age 15-49) data for Kaduna State**

```{r data}
data <- read.csv("Data/kaduna_health_sub.csv")
r <- raster("Data/nga_pop_wocba.tif")
boundaries <- shapefile("Data/kaduna_wards.shp")
```

We will view our health facilty data below. In total there are `r nrow(data)` health centres recorded within the dataset. This includes data such as their location, whether the type of health centre *(primary, secondary, terariary)*, and whether it is private or publicly owned. An example of some of the data is shown below:

```{r view}
head(data[8:18], n = 3)
```

For our example, we are only interested in **public** health centers. We will therefore filter the dataset below:

```{r subset}
data_sub <- data[data$ownership == "Public", ]
```

In order to conduct spatial analysis on the data, we must convert these into a spatial object. 

```{r sp}
points <- SpatialPointsDataFrame(data_sub[, c("longitude", "latitude")], data_sub)
```

We will quickly visualise this data below:

```{r locations, message=FALSE}
theme <- theme(
  axis.ticks.y = element_blank(), axis.text.y = element_blank(),
  axis.ticks.x = element_blank(), axis.text.x = element_blank()
)

ggplot() + 
  geom_polygon(
  data = boundaries, aes(x = long, y = lat, group = group), fill = "white", colour = "black", alpha = 1) +
  geom_point(data = data_sub, aes(x = longitude, y = latitude), color = "blue") +
  labs(x = "", y = "") +
  theme
```

Visualise the gridded population data:

```{r grid vis}
plot(boundaries)
plot(r, add = TRUE)
```

# Coverage Calculations

We are interested in X...

Create a set of voronoi polygons from the point locations. We can use the `voronoi` function from the package **dismo** to calculate these:

```{r voronois}
v <- voronoi(points)
```

Again, we will visualise these below:

```{r, message=FALSE}
ggplot() + geom_polygon(
  data = v, aes(x = long, y = lat, group = group), fill = "white",
  colour = "black", alpha = 1
) +
  labs(x = "", y = "") +
  theme +
  geom_point(data = data_sub, aes(x = longitude, y = latitude), color = "blue")
```

This shows an estimation of health facility coverage. Based on area, you can start to identify places that could benefit from new facilities. However, we first need to take the population estimates into account to find out how many people are being covered. Let's compare population density to health facility density.

<!--- It would be nice to explain why you are using a focal window --->

We are going to sum each in a 3x3 window to look at population density and health facility density.

```{r pop dens, error=TRUE}
a <- focalWeight(r, 0.025, "circle")
a[a > 0] <- 1
p_dens <- focal(r, w = a, fun = sum, na.rm = TRUE)
plot(p_dens, main = "Population Density")

rast <- rasterize(points, r, field = 1)
hf_dens <- focal(rast, w = a, fun = sum, na.rm = TRUE)
plot(hf_dens, main = "Health Facility Density")
```

# Identifying Relationships

We are going to look at the relationship between the facility density and population density. We can also use this to derive targets of the number of people per health facility.

```{r lm,warning=FALSE,message=FALSE}
plot(p_dens, hf_dens, xlab = "population density 3km", ylab = "hfac density 3km")
abline(lm(values(hf_dens) ~ values(p_dens)))
```

We could use a target of 5000 people per health facility to look at areas where health facilities are going to be serving too many people.Using zonal statistics, we need to calculate the number of people in each voronoi polyon. 

```{r zonal,message=FALSE}
z.stat <- zonal.stats(v, r, stat = sum, trace = FALSE, plot = FALSE)
z <- data.frame(ID = v$name, sum = z.stat)
head(z, nrow = 3)

v@data <- data.frame(v@data, z[match(v@data[, "name"], z[, "ID"]), ])
```

Here, we are subsetting the data to only include those voronois above the 'target' that can be seen in blue. These would be recommended for further intervention, such as the expansion of existing health facilities or placement of new facilities.

```{r target,message=FALSE}
sub <- subset(v, sum > 5000)

ggplot() + geom_polygon(
  data = v, aes(x = long, y = lat, group = group), fill = "white",
  colour = "black", alpha = 1
) +
  geom_polygon(
    data = sub, aes(x = long, y = lat, group = group), fill = "skyblue2",
    colour = "blue", alpha = 1
  ) +
  theme +
  labs(x = "", y = "")
```

# Further examples


# Further Reading




# References